Cultural Transmission Can Explain the Evolution of Cooperative Behavior

### 2020-07-26 ms/main.tex
merged result 3 and 4; separated appendix B to B and C; using appendix package; appendix eqs numbers; fixed sympy citation key

\usepackage{amssymb,amsmath,amsthm}
\usepackage{commath}
\usepackage[hyphens]{url}
\usepackage[dvipsnames]{xcolor}
\usepackage[unicode=true,colorlinks=true,urlcolor=CadetBlue,citecolor=black,linkcolor=black]{hyperref}
\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
\usepackage{authblk}
\usepackage{longtable}
\usepackage{multirow}
\usepackage{booktabs}

@@

@@

@@

% Introduction
\section*{Introduction}
Cooperative behavior can harm an individual's fitness and increase the fitness of its conspecifics or competitors~\citep{axelrod1981evolution}.
Nevertheless, cooperative behavior appears to occur in many non-human animals~\citep{dugatkin1997cooperation}, for example rats~\citep{rice1962altruism} and birds~\citep{krams2008experimental}.
Evolution of cooperative behavior remains  an important  conundrum in evolutionary biology.
\\
Cultural evolution is an evolutionary theory of individual and group-level change in beliefs, behaviors, and norms \citep{cavalli1981cultural}.
Culture has significant impact on the behavior of humans~\citep{ihara2004cultural,jeong2018bronze} as well as non-human animals~\citep{bonner2018evolution}.
Cultural transmission allows an individual to acquire attitudes and behavioral traits from other individuals in its social group through imitation, learning, or other modes of communication \citep{cavalli1981cultural,richerson2008not}.
Here we attempt to determine to what extent cultural transmission can explain the evolution of cooperative behavior.

% subsection
%\subsection*{Theories for evolution of cooperation}
Three major theories have been proposed to explain the evolution of cooperative behavior:\\
1. \emph{Kin selection} theory posits that natural selection can favor cooperation between related individuals. The importance of relatedness to the evolution of cooperation and altruism was shown by~\citet{hamilton1964genetical}. According to Hamilton,  an allele that determines altruistic behavior should increase in frequency when  the reproductive cost to the actor that cooperates, $c$, is less than the benefit to the recipient, $b$, times the genetic relatedness between the recipient and the actor, $r$.
This is also known as Hamilton's Rule:
\begin{equation} \label{eq:hamilton_rule}
c < b \cdot r.
\end{equation}

@@

Group selection has been criticized by biologists who advocate a gene-centered view of evolution. It has also been criticized because for cooperation to take over the population it must have higher fitness than defection, while under group selection theory the fitness of cooperators at the individual level is lower than the fitness of defectors. Thus a trait with a lower fitness taking over the population is a contradiction %TODO ref.
\citet{eldakar2011eight} reject this argument, claiming that it is a tautology and does not qualify as an argument against group selection. The distinction between individual and group selection requires a comparison of fitness differentials within and between groups in a multi-group population, and when a trait  evolves by group selection, despite having lower fitness within a group, that group might have higher average fitness in competition with other groups, all things considered. % TODO cite Simpsons' paradox?

The above theories assume that cooperation is genetically determined, which raises the question: \emph{Is it possible that cooperation is determined by non-genetic factors?}
\citet{feldman1985gene} introduced the first model for the evolution of altruism by cultural transmission. They showed that under vertical (parent-to-offspring) cultural transmission, Hamilton's rule does not govern the evolution of parent-to-offspring or sib-to-sib altruism.
Non-vertical transmission may be either a horizontal or oblique: horizontal transmission occurs between individuals from the same generation, while oblique transmission occurs from an adult to an unrelated offspring. 
Evolution under either of these transmission models  can be be more rapid than under pure vertical transmission~\citep{cavalli1981cultural,ram2018evolution}.

\citet{lewin2017microbes} have demonstrated that \emph{non-vertical transmission}, mediated by microbes that manipulate their host behavior, can help to explain the evolution of cooperative behavior. Interestingly, some of the analysis made by \citet{lewin2017microbes} can be applied to cultural transmission, because models of cultural transmission are mathematically similar to those for transmission of infectious diseases~\citep{cavalli1981cultural}.

Here we hypothesize that non-vertical cultural transmission can explain the evolution of cooperation. 
To test this hypothesis, we suggest a model in which behavioral changes are mediated by cultural transmission that can occur during social interaction. For example, if an individual interacts with a cooperative individual, it might learn that cooperation is a positive behavior and will cooperate in the future. 
We develop cultural evolution models that include both vertical and non-vertical transmission of cooperation and investigate these models using both mathematical analysis and simulations.  
Our results demonstrate the conditions in which cooperation can evolve when it is culturally, rather than genetically, transmitted.

% Models and Methods
\section*{Models and Methods}

First, we focus on the evolution of cooperation in a fully mixed population where cooperation is modeled using the prisoner's dilemma. % yr: add ref

@@

Individuals interact according to a prisoner's dilemma.
Specifically, individuals interact in pairs; a cooperator suffers a fitness cost $0<c<1$, and its partner gains a fitness benefit $b$, where we assume $b>c$. \autoref{table:prisoner_payoff} shows the payoff matrix, i.e. the fitness, of an individual with phenotype $\phi_1$ when interacting with a partner of phenotype $\phi_2$ (we assume $b>c>0$):

% Table: payoff matrix

\begin{table}[h]
\centering
\begin{tabular}{lll}
\toprule
           & $\phi_2=A$ & $\phi_2=B$ \\ \cmidrule(r){1-3}
$\phi_1=A$ & $1+b-c$ & $1-c$ \\
$\phi_1=B$ & $1+b$   & $1$
\\ \bottomrule
\end{tabular}
\caption{\textbf{Payoff matrix for prisoner's dilemma.}
The fitness of phenotype $\phi_1$ when interacting with phenotype $\phi_2$. $A$ is a cooperative phenotype, $B$ is a defector phenotype, $b$ is the benefit gained by an individual interacting with a cooperator, and $c$ is the cost of cooperation. $b>c>0$.
}
\label{table:prisoner_payoff}
\end{table}

@@

Horizontal cultural transmission occurs between peers. 
It may occur between social partners with probability $\alpha$, or between a random pair with probability $1-\alpha$ (see~\textbf{\autoref{fig:horizontal}}).
Horizontal transmission is not always successful, as one peer may reject the other's phenotype. The probability for successful transmission of phenotypes $A$ and $B$ are $T_A$ and $T_B$, respectively.\\
\autoref{table:interactions} contains the probability of first interactor to be A following interaction by interaction type.
Therefore, the frequency $p'$ of phenotype $A$ among adults in the next generation, after horizontal transmission, is 
\begin{equation}\label{eq:nextgen_adults}
\begin{aligned}
p'
& = \hat{p}^2 [\alpha + (1-\alpha)(\hat{p} + (1-\hat{p})(1-T_B))] \\
& + \hat{p}(1-\hat{p}) [\alpha(1-T_B) + (1-\alpha)(\hat{p} + (1-\hat{p})(1-T_B))] \\
& + (1-\hat{p})\hat{p} [\alpha T_A + (1-\alpha) \hat{p} T_A ] \\
& + (1-\hat{p})^2 [(1-\alpha) \hat{p} T_A] \;,
\end{aligned}
\end{equation}
which simplifies to
\begin{equation}\label{eq:nextgen_adults_slimpify}
p' = \hat{p}^2(T_B-T_A) + \hat{p}(1+T_A-T_B) .
\end{equation}

The frequency of $A$ among parents (after selection) follows a similar dynamic, but also includes the effect of natural selection, and is therefore
\begin{equation}\label{eq:nextgen_parents}
\begin{aligned}
\bar{w} \tilde{p}'
& = \hat{p}^2 (1+b-c) [\alpha + (1-\alpha)(\hat{p} + (1-\hat{p})(1-T_B))] \\
& + \hat{p}(1-\hat{p}) (1-c) [\alpha(1-T_B) + (1-\alpha)(\hat{p} + (1-\hat{p})(1-T_B))] \\
& + (1-\hat{p})\hat{p} (1+b) [\alpha T_A + (1-\alpha) \hat{p} T_A ] \\
& + (1-\hat{p})^2 [(1-\alpha) \hat{p} T_A] \;,
\end{aligned}
\end{equation}
where fitness values are taken from \autoref{table:interactions}, and 
\begin{equation} \label{eq:mean_fitness}
\bar{w} =  1 + \hat{p}(b-c)
\end{equation}
is the population mean fitness.
Eq~\ref{eq:nextgen_parents} can be simplified to
\begin{equation}\label{eq:nextgen_parents_simplified}
\begin{aligned}
\bar{w} \tilde{p}'
& = \hat{p}^2 (1+b-c) \big(1-(1-\hat{p})(1-\alpha)T_B)\big) \\
& + \hat{p}(1-\hat{p}) (1-c) \big(\hat{p}(1-\alpha)T_B+1-T_B\big) \\
& + (1-\hat{p})\hat{p} (1+b) \big(\hat{p}(1-\alpha) + \alpha\big) T_A \\
& + (1-\hat{p})^2 \hat{p} (1-\alpha) T_A \;.
\end{aligned}
\end{equation}

% Table: interactions
\begin{table}[]
\begin{tabular}{@{}llllll@{}}
\toprule
\multirow{2}{*}{Phenotype $\phi_1$} &
  \multirow{2}{*}{Phenotype $\phi_2$} &
  \multirow{2}{*}{Frequency} &
  \multirow{2}{*}{Fitness of $\phi_1$} &
  \multicolumn{2}{l}{$P(\phi_1=A)$ via horizontal transmission:} \\ \cmidrule(l){5-6} 
    &     &                      &         & from partner, $\alpha$ & from population, $(1-\alpha)$ \\ \cmidrule(r){1-6}
$A$ & $A$ & $\hat{p}^2$          & $1+b-c$ & 1                      & $\hat{p}+(1-\hat{p})(1-T_B)$  \\
$A$ & $B$ & $\hat{p}(1-\hat{p})$ & $1-c$   & $1-T_B$                & $\hat{p}+(1-\hat{p})(1-T_B)$  \\
$B$ & $A$ & $\hat{p}(1-\hat{p})$ & $1+b$   & $T_A$                  & $\hat{p} T_A$                 \\
$B$ & $B$ & $(1-\hat{p})^2$      & $1$     & $0$                    & $\hat{p} T_A$                 \\ \bottomrule
\end{tabular}
\caption{\textbf{Interaction frequency, fitness, and transmission probabilities.}}
\label{table:interactions}
\end{table}

% Figure: model illustration
\begin{figure}[b]
  \centering
  \includegraphics[scale=1]{figure1.pdf}
  \caption{\textbf{Cultural horizontal transmission.} Transmission occurs between interacting partners with probability $\alpha$ (left) or between two random peers with probability $1-\alpha$.}
  \label{fig:horizontal}
\end{figure}

% Results
\section*{Results}

We start by studying specific cases, for which we can derive general results, and then 
we use numerical simulation to analyze more complex cases.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Without Oblique Transmission}

With only vertical and horizontal transmission, i.e. $v=1$, eq.~\ref{eq:horizontal} becomes
$\hat{p} =  \tilde{p}$,
and eq.~\ref{eq:nextgen_parents_simplified} for the change in frequency $p'$ of phenotype $A$ among parents can be written as

\begin{equation} 
\begin{split}\label{eq:nextgen_parents_vertical_only}
\bar{w} \tilde{p}' 
& = \tilde{p}^2 (1+b-c) [1 - (1-\tilde{p}) (1-\alpha) T_B] \\
& + \tilde{p}(1-\tilde{p}) (1-c) [\tilde{p} (1-\alpha) T_B + 1 - T_B] \\
& + \tilde{p}(1-\tilde{p}) (1+b) [\tilde{p} (1-\alpha) + \alpha] T_A \\
& + (1-\tilde{p})^2 \tilde{p} (1-\alpha) T_A
\end{split}
\end{equation}

@@

\textbf{Result 1: Vertical and horizontal transmission of cooperation.}
From eq.~\ref{eq:nextgen_parents_vertical_only}, if
\begin{equation} \label{eq:unequal_transmission}
c \cdot (1-T_B) < b \cdot  \alpha T_A  + (T_A - T_B) [1 + b\tilde{p}(1-\alpha)] ,
\end{equation}
then $\tilde{p}' > \tilde{p}$, and the frequency of the cooperator phenotype $A$ among parents increases every generation.

\textbf{Corollary 1.1: Symmetric horizontal transmission.}
If $T=T_A=T_B$, then
\begin{equation}
\label{eq:equal_transmission}
c < b \cdot \frac{\alpha T}{1-T},
\end{equation}
which can be seen as a cultural version of \emph{Hamilton's rule} (eq.~\ref{eq:hamilton_rule}), where $r_H=\alpha T/(1-T)$ can be thought of as \emph{horizontal relatedness}.
Therefore, if the cost $c$ is less than the benefit $b$ times the horizontal relatedness $r_H$, then cooperation will take over of the population (see~\textbf{\autoref{fig:results_a}}).

@@

In this case $\alpha=1$, and horizontal transmission can only occur as a result of cooperative interactions.
Therefore, eq.~\ref{eq:unequal_transmission}, which determines the conditions for evolution of cooperation, becomes
\begin{equation}\label{eq:horizontal_transmission_only_after_interaction}
c \cdot (1-T_B) < b \cdot T_A + (T_A - T_B).
\end{equation}
This is equivalent to a result by~\citet[eq.~1]{lewin2017microbes}.

Eq.~\ref{eq:horizontal_transmission_only_after_interaction} can be written as
\begin{equation} \label{eq:horizontal_transmission_only_after_interaction2}
1 - (1-T_B)(1-c) < T_A \cdot (1+b) ,
\end{equation}
which provides an interesting interpretation for the success of cooperation. 
Consider an interaction between two individuals: a cooperator and a defector.
$(1-T_B)(1-c)$ is the probability that the cooperator remains cooperative and also reproduces. 
Therefore, $1 - (1-T_B)(1-c)$ is the probability that either the cooperator becomes a defector, \emph{or} that it fails to reproduce.
This is the effective cost for cooperation from this interaction.
$T_A\cdot(1+b)$ is the probability that the defector becomes cooperative and reproduces.
This is the effective benefit for cooperation from this interaction.
So, eq.~\ref{eq:horizontal_transmission_only_after_interaction2} means that cooperation can evolve if effective cost is less than the effective benefit.

@@

In this case $\alpha=0$, and horizontal transmission is entirely independent from cooperative interactions.  
Then, eq.~\ref{eq:unequal_transmission} becomes
\begin{equation} \label{eq:horizontal_transmission_from_population}
c\cdot(1-T_B) < b \cdot \tilde{p} \;(T_A-T_B) + (T_A-T_B). 
\end{equation}

@@

With only oblique and horizontal transmission, i.e. $v = 0$, eq.~\ref{eq:horizontal} becomes $\hat{p}=p$ and eq.~\ref{eq:nextgen_adults_slimpify} becomes % TODO check this, I changed the ref from nextgen_parents_simplified to nextgen_adults_slimpify
\begin{equation}  \label{eq:nextgen_parents_oblique_only}
p' = p^2 (T_B-T_A) + p (1+T_A-T_B) ,
\end{equation}

@@

In this case $0<v<1$, and the recursion system is more complex.
Therefore, we focus on local, rather than global, stability.
To proceed, we note that 
eq.~\ref{eq:horizontal} can give $\hat{p}'$ as a function of both $p'$ and $\tilde{p}'$,
eq.~\ref{eq:nextgen_adults_slimpify} gives $p'$ as a function of $\hat{p}$, and 
eq.~\ref{eq:nextgen_parents_simplified} gives $\tilde{p}'$ as a function of $\hat{p}$. 
Combining these equations, we find an equation for $\hat{p}'$ as a function of $\hat{p}$.
We then determine the equilibria, which are solutions of $\hat{p}' = \hat{p}$, and analyse their local stability: an equilibrium $\hat{p}^*$ is locally stable when the derivative of $f(\hat{p})$ at the equilibrium is negative, $f'(\hat{p}^*)<0$. %MWF: Define f(\hat p).

We start with the simple case of symmetrical horizontal transmission, $T=T_A=T_B$ and 
%MWF: Write the full recursion.
apply eqs.~\ref{eq:horizontal}, \ref{eq:nextgen_adults_slimpify}, and \ref{eq:nextgen_parents_simplified} to obtain 
\begin{equation} \label{eq:equal_horizontal_transmission}
\begin{aligned}
  %\hat{p}' - \hat{p} = 
  %\hat{p} ^2(-\alpha bvT + cv(1-T)) + \hat{p} (\alpha bvT -cv(1-T)) = 
  f(\hat{p}) &= 
  \bar{w}(\hat{p}' - \hat{p}) = \\
  &\hat{p}(1-\hat{p})\big[\alpha bvT - cv(1-T)\big]. % shorter version
\end{aligned}
\end{equation}

The equilibria are solutions of $f(\hat{p})=0$, or $\hat{p}' = \hat{p}$.
It is easy to verify that fixation of either phenotype, $\hat{p} =  0$ and $\hat{p} = 1$, is an equilibrium.
Since the derivative of $f'(\hat{p}$ is 
\begin{equation}
f'(\hat{p})=(1-2\hat{p})\big[\alpha bvT - cv(1-T)\big],
\end{equation}
then the condition for local stability of $\hat{p}=1$ is
\begin{equation} \label{eq:derivative_of_phattag-phat}
	f'(1) =
% 2\cdot\hat{p} (-\alpha bvT + cv(1-T)) + (\alpha bvT -cv(1-T)).
%	(1- 2\hat{p}) (\alpha bvT - cv(1-T)) \mid_{\hat{p}=1} = 
	-\alpha bvT + cv(1-T) < 0,
\end{equation}
which produces the following result.

>>>>

\usepackage{amssymb,amsmath,amsthm}
\usepackage{commath}
\usepackage[hyphens]{url}
\usepackage[dvipsnames]{xcolor}
\usepackage[unicode=true,colorlinks=true,urlcolor=CadetBlue,citecolor=black,linkcolor=black]{hyperref}
\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
\usepackage{authblk}
\usepackage{longtable}
\usepackage{multirow}
\usepackage{booktabs}
\usepackage{lipsum}  
\usepackage[title,page]{appendix}
\usepackage{chngcntr}

@@

% line numbers
 \usepackage[displaymath, mathlines]{lineno}
 \renewcommand\linenumberfont{\normalfont\small\sffamily}
\linenumbers
\modulolinenumbers[2]

@@

% Supplementary
% https://support.authorea.com/en-us/article/how-to-create-an-appendix-section-or-supplementary-information-1g25i5a/
\newcommand{\beginsupplement}{%
      	\setcounter{table}{0}
        \renewcommand{\thetable}{S\arabic{table}}%
        \setcounter{figure}{0}
        \renewcommand{\thefigure}{S\arabic{figure}}%
		\setcounter{equation}{0}
        \renewcommand{\theequation}{A\arabic{equation}}%
}

@@

%%% Introduction
\section*{Introduction}
Cooperative behavior can harm an individual's fitness and increase the fitness of its conspecifics or competitors~\citep{axelrod1981evolution}.
Nevertheless, cooperative behavior appears to occur in many non-human animals~\citep{dugatkin1997cooperation}, for example rats~\citep{rice1962altruism} and birds~\citep{krams2008experimental}.
Evolution of cooperative behavior remains an important conundrum in evolutionary biology.
Three major theories have been proposed to explain the evolution of cooperative behavior:

1. \emph{Kin selection} theory posits that natural selection can favor cooperation between related individuals. The importance of relatedness to the evolution of cooperation and altruism was shown by~\citet{hamilton1964genetical}. According to Hamilton,  an allele that determines altruistic behavior should increase in frequency when  the reproductive cost to the actor that cooperates, $c$, is less than the benefit to the recipient, $b$, times the genetic relatedness between the recipient and the actor, $r$.
This is also known as Hamilton's Rule:
\begin{equation} \label{eq:hamilton_rule}
c < b \cdot r.
\end{equation}

@@

Group selection has been criticized by biologists who advocate a gene-centered view of evolution. % TODO ref
It has also been criticized because for cooperation to take over the population it must have higher fitness than defection, while under group selection theory the fitness of cooperators at the individual level is lower than the fitness of defectors. Thus a trait with a lower fitness taking over the population is a contradiction %TODO ref.
\citet{eldakar2011eight} reject this argument, claiming that it is a tautology and does not qualify as an argument against group selection. The distinction between individual and group selection requires a comparison of fitness differentials within and between groups in a multi-group population, and when a trait  evolves by group selection, despite having lower fitness within a group, that group might have higher average fitness in competition with other groups, all things considered. % TODO cite Simpsons' paradox?

These theories assume that cooperation is genetically determined, which raises the question: \emph{Is it possible that cooperation is determined by non-genetic factors?}
Culture has significant impact on the behavior of humans~\citep{ihara2004cultural,jeong2018bronze} as well as non-human animals~\citep{bonner2018evolution}.
Here we attempt to determine to what extent the evolution of cooperative behavior can be explained by \emph{cultural transmission},
which allows an individual to acquire attitudes and behavioral traits from other individuals in its social group through imitation, learning, or other modes of communication \citep{cavalli1981cultural,richerson2008not}.
\citet{feldman1985gene} introduced the first model for the evolution of altruism by cultural transmission.
They showed that under vertical (parent-to-offspring) cultural transmission, Hamilton's rule does not govern the evolution of parent-to-offspring or sib-to-sib altruism.

\emph{Non-vertical transmission} may be either horizontal or oblique: horizontal transmission occurs between individuals from the same generation, while oblique transmission occurs from adults to unrelated offspring. 
Evolution under either of these transmission models  can be be more rapid than under pure vertical transmission~\citep{cavalli1981cultural,ram2018evolution}.
\citet{lewin2017microbes} have demonstrated that non-vertical transmission, mediated by microbes that manipulate their host behavior, can help to explain the evolution of cooperative behavior. Interestingly, some of their analysis can be applied to cultural transmission, because models of cultural transmission are mathematically similar to those for transmission of infectious diseases~\citep{cavalli1981cultural}.

Here we hypothesize that non-vertical cultural transmission can explain the evolution of cooperation. 
To test this hypothesis, we suggest a model in which behavioral changes are mediated by cultural transmission that can occur during social interactions. For example, if an individual interacts with a cooperative individual, it might learn that cooperation is a positive behavior and will be cooperative in the future. 
We develop cultural evolution models that include both vertical and non-vertical transmission of cooperation and investigate these models using mathematical analysis and simulations.  
Our results demonstrate cultural transmission can facilitate the evolution of cooperation even when genetic transmission cannot.
These results suggest that further research on the evolution of cooperation should account for non-vertical transmission and that treatment of cooperation as a cultural, rather then genetic trait, can lead to a better understanding of this important and enigmatic phenomenon.

% Models
\section*{Models}

We focus on the evolution of cooperation in a fully mixed population where cooperation is modeled using the \emph{prisoner's dilemma}. % yr: add ref

@@

Individuals interact according to a prisoner's dilemma.
Specifically, individuals interact in pairs; a cooperator suffers a fitness cost $0<c<1$, and its partner gains a fitness benefit $b$, where we assume $b>c>0$. \textbf{\autoref{table:prisoner_payoff}} shows the payoff matrix, i.e. the fitness of an individual with phenotype $\phi_1$ when interacting with a partner of phenotype $\phi_2$.

%%% Table: payoff matrix
\begin{table}[h]
\centering
\begin{tabular}{lll}
\toprule
           & $\phi_2=A$ & $\phi_2=B$ \\ \cmidrule(r){1-3}
$\phi_1=A$ & $1+b-c$ & $1-c$ \\
$\phi_1=B$ & $1+b$   & $1$
\\ \bottomrule
\end{tabular}
\caption{\textbf{Payoff matrix for prisoner's dilemma.}
The fitness of phenotype $\phi_1$ when interacting with phenotype $\phi_2$. $A$ is a cooperative phenotype, $B$ is a defector phenotype, $b$ is the benefit gained by an individual interacting with a cooperator, and $c$ is the cost of cooperation. $b>c>0$.
}
\label{table:prisoner_payoff}
\end{table}
%%%

@@

Horizontal cultural transmission occurs between peers. 
It may occur between social partners with probability $\alpha$, or between a random pair with probability $1-\alpha$ (see~\textbf{\autoref{fig:horizontal}}).
Horizontal transmission is not always successful, as one peer may reject the other's phenotype. The probability for successful horizontal transmission of phenotypes $A$ and $B$ are $T_A$ and $T_B$, respectively (\textbf{\autoref{table:interactions}}).

Therefore, the frequency $p'$ of phenotype $A$ among adults in the next generation, after horizontal transmission, is 
\begin{equation}\label{eq:nextgen_adults}
\begin{aligned}
p'
& = \hat{p}^2 [\alpha + (1-\alpha)(\hat{p} + (1-\hat{p})(1-T_B))] \\
& + \hat{p}(1-\hat{p}) [\alpha(1-T_B) + (1-\alpha)(\hat{p} + (1-\hat{p})(1-T_B))] \\
& + (1-\hat{p})\hat{p} [\alpha T_A + (1-\alpha) \hat{p} T_A ] \\
& + (1-\hat{p})^2 [(1-\alpha) \hat{p} T_A] \;,
\end{aligned}
\end{equation}
which simplifies to
\begin{equation}\label{eq:nextgen_adults_slimpify}
p' = \hat{p}^2(T_B-T_A) + \hat{p}(1+T_A-T_B) .
\end{equation}

The frequency of $A$ among parents (i.e. after selection) follows a similar dynamic, but also includes the effect of natural selection, and is therefore
\begin{equation}\label{eq:nextgen_parents}
\begin{aligned}
\bar{w} \tilde{p}'
& = \hat{p}^2 (1+b-c) [\alpha + (1-\alpha)(\hat{p} + (1-\hat{p})(1-T_B))] \\
& + \hat{p}(1-\hat{p}) (1-c) [\alpha(1-T_B) + (1-\alpha)(\hat{p} + (1-\hat{p})(1-T_B))] \\
& + (1-\hat{p})\hat{p} (1+b) [\alpha T_A + (1-\alpha) \hat{p} T_A ] \\
& + (1-\hat{p})^2 [(1-\alpha) \hat{p} T_A] \;,
\end{aligned}
\end{equation}
where fitness values are taken from \textbf{\autoref{table:prisoner_payoff}} and \textbf{\autoref{table:interactions}}, and the population mean fitness is
\begin{equation} \label{eq:mean_fitness}
\bar{w} =  1 + \hat{p}(b-c).
\end{equation}

\autoref{eq:nextgen_parents} can be simplified to
\begin{equation}\label{eq:nextgen_parents_simplified}
\begin{aligned}
\bar{w} \tilde{p}'
& = \hat{p}^2 (1+b-c) \big(1-(1-\hat{p})(1-\alpha)T_B)\big) \\
& + \hat{p}(1-\hat{p}) (1-c) \big(\hat{p}(1-\alpha)T_B+1-T_B\big) \\
& + (1-\hat{p})\hat{p} (1+b) \big(\hat{p}(1-\alpha) + \alpha\big) T_A \\
& + (1-\hat{p})^2 \hat{p} (1-\alpha) T_A \;.
\end{aligned}
\end{equation}

%%% Table: interactions
\begin{table}[]
\begin{tabular}{@{}llllll@{}}
\toprule
\multirow{2}{*}{Phenotype $\phi_1$} &
  \multirow{2}{*}{Phenotype $\phi_2$} &
  \multirow{2}{*}{Frequency} &
  \multirow{2}{*}{Fitness of $\phi_1$} &
  \multicolumn{2}{l}{$P(\phi_1=A)$ via horizontal transmission:} \\ \cmidrule(l){5-6} 
    &     &                      &         & from partner, $\alpha$ & from population, $(1-\alpha)$ \\ \cmidrule(r){1-6}
$A$ & $A$ & $\hat{p}^2$          & $1+b-c$ & 1                      & $\hat{p}+(1-\hat{p})(1-T_B)$  \\
$A$ & $B$ & $\hat{p}(1-\hat{p})$ & $1-c$   & $1-T_B$                & $\hat{p}+(1-\hat{p})(1-T_B)$  \\
$B$ & $A$ & $\hat{p}(1-\hat{p})$ & $1+b$   & $T_A$                  & $\hat{p} T_A$                 \\
$B$ & $B$ & $(1-\hat{p})^2$      & $1$     & $0$                    & $\hat{p} T_A$                 \\ \bottomrule
\end{tabular}
\caption{\textbf{Interaction frequency, fitness, and transmission probabilities.}}
\label{table:interactions}
\end{table}
%%%

%%% Figure: model illustration
\begin{figure}[b]
  \centering
  \includegraphics[scale=1]{figure1.pdf}
  \caption{\textbf{Cultural horizontal transmission.} Transmission occurs between interacting partners with probability $\alpha$ (left) or between two random peers with probability $1-\alpha$.}
  \label{fig:horizontal}
\end{figure}
%%%

%%% Results
\section*{Results}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Without Oblique Transmission}

With only vertical and horizontal transmission, i.e. $v=1$, \autoref{eq:horizontal} becomes
$\hat{p} =  \tilde{p}$,
and \autoref{eq:nextgen_parents_simplified} for the change in frequency $p'$ of phenotype $A$ among parents can be written as
\begin{equation} \label{eq:nextgen_parents_vertical_only} 
\begin{aligned}
\bar{w} \tilde{p}' 
& = \tilde{p}^2 (1+b-c) [1 - (1-\tilde{p}) (1-\alpha) T_B] \\
& + \tilde{p}(1-\tilde{p}) (1-c) [\tilde{p} (1-\alpha) T_B + 1 - T_B] \\
& + \tilde{p}(1-\tilde{p}) (1+b) [\tilde{p} (1-\alpha) + \alpha] T_A \\
& + (1-\tilde{p})^2 \tilde{p} (1-\alpha) T_A .
\end{aligned}
\end{equation}

@@

\textbf{Result 1: Vertical and horizontal transmission of cooperation.}
From \autoref{eq:nextgen_parents_vertical_only}, if
\begin{equation} \label{eq:unequal_transmission}
c \cdot (1-T_B) < b \cdot  \alpha T_A  + (T_A - T_B) [1 + b\tilde{p}(1-\alpha)] ,
\end{equation}
then $\tilde{p}' > \tilde{p}$, and the frequency of the cooperator phenotype $A$ among parents increases every generation.

\textbf{Corollary 1.1: Symmetric horizontal transmission.}
If $T=T_A=T_B$, then
\begin{equation}
\label{eq:equal_transmission}
c < b \cdot \frac{\alpha T}{1-T},
\end{equation}
which can be seen as a cultural version of \emph{Hamilton's rule} (\autoref{eq:hamilton_rule}), where $r_H=\alpha T/(1-T)$ can be thought of as \emph{horizontal relatedness}.
Therefore, if the cost $c$ is less than the benefit $b$ times the horizontal relatedness $r_H$, then cooperation will take over of the population (see~\textbf{\autoref{fig:results_a}}).

@@

In this case $\alpha=1$, and horizontal transmission can only occur as a result of cooperative interactions.
Therefore, \autoref{eq:unequal_transmission}, which determines the conditions for evolution of cooperation, becomes
\begin{equation}\label{eq:horizontal_transmission_only_after_interaction}
c \cdot (1-T_B) < b \cdot T_A + (T_A - T_B).
\end{equation}
This is equivalent to a result by~\citet[eq.~1]{lewin2017microbes}.

\autoref{eq:horizontal_transmission_only_after_interaction} can be written as
\begin{equation} \label{eq:horizontal_transmission_only_after_interaction2}
1 - (1-T_B)(1-c) < T_A \cdot (1+b) ,
\end{equation}
which provides an interesting interpretation for the success of cooperation. 
Consider an interaction between two individuals: a cooperator and a defector.
$(1-T_B)(1-c)$ is the probability that the cooperator remains cooperative and also reproduces. 
Therefore, $1 - (1-T_B)(1-c)$ is the probability that either the cooperator becomes a defector, \emph{or} that it fails to reproduce.
This is the effective cost for cooperation from this interaction.
$T_A\cdot(1+b)$ is the probability that the defector becomes cooperative and reproduces.
This is the effective benefit for cooperation from this interaction.
So, \autoref{eq:horizontal_transmission_only_after_interaction2} means that cooperation can evolve if the effective cost is less than the effective benefit.

@@

In this case $\alpha=0$, and horizontal transmission is entirely independent from cooperative interactions.  
Then, \autoref{eq:unequal_transmission} becomes
\begin{equation} \label{eq:horizontal_transmission_from_population}
c\cdot(1-T_B) < b \cdot \tilde{p} \;(T_A-T_B) + (T_A-T_B). 
\end{equation}

@@

With only oblique and horizontal transmission, i.e. $v = 0$, \autoref{eq:horizontal} becomes $\hat{p}=p$ and \autoref{eq:nextgen_adults_slimpify} becomes % TODO check this, I changed the ref from nextgen_parents_simplified to nextgen_adults_slimpify
\begin{equation}  \label{eq:nextgen_parents_oblique_only}
p' = p^2 (T_B-T_A) + p (1+T_A-T_B) ,
\end{equation}

@@

In this case $0<v<1$, and the recursion system is more complex.
Therefore, we focus on local stability, rather than global stability.
To proceed, we note that 
\autoref{eq:horizontal} can give $\hat{p}'$ as a function of both $p'$ and $\tilde{p}'$,
\autoref{eq:nextgen_adults_slimpify} gives $p'$ as a function of $\tilde{p}$, and 
\autoref{eq:nextgen_parents_simplified} gives $\tilde{p}'$ as a function of $\hat{p}$. 
Combining these equations, we find an equation for $\hat{p}'$ as a function of $\hat{p}$, see Appendix~\autoref{sec:appendixB}.
We then determine the equilibria, which are solutions of $\hat{p}' = \hat{p}$, and analyse their local stability.
%: an equilibrium $\hat{p}^*$ is locally stable when the derivative of $f(\hat{p})=\bar{w}(\hat{p}'-\hat{p})$ at the equilibrium is negative, $f'(\hat{p}^*)<0$.

%We start with the simple case of symmetrical horizontal transmission, $T=T_A=T_B$ and apply \autoref{eq:horizontal}, \autoref{eq:nextgen_adults_slimpify}, and \autoref{eq:nextgen_parents_simplified} to obtain 
%\begin{equation} \label{eq:equal_horizontal_transmission}
%\begin{aligned}
%  f(\hat{p}) &= 
%  \bar{w}(\hat{p}' - \hat{p}) = \\
%  &\hat{p}(1-\hat{p})\big[\alpha bvT - cv(1-T)\big].\end{aligned}
%\end{equation}
%A detailed explain of how we get eq.~\ref{eq:equal_horizontal_transmission} can be found in appendix B.
%The equilibria are solutions of $f(\hat{p})=0$, or $\hat{p}' = \hat{p}$.
%It is easy to verify that fixation of either phenotype, $\hat{p} =  0$ and $\hat{p} = 1$, is an equilibrium.
%Since the derivative of $f'(\hat{p}$ is 
%\begin{equation}
%f'(\hat{p})=(1-2\hat{p})\big[\alpha bvT - cv(1-T)\big],
%\end{equation}
%then the condition for local stability of $\hat{p}=1$ is
%\begin{equation} \label{eq:derivative_of_phattag-phat}
%	f'(1) =	-\alpha bvT + cv(1-T) < 0,
%\end{equation}
%which gives the following result.
%
%\textbf{Result 3: Oblique and vertical transmission with symmetric horizontal transmission.}
%If horizontal transmission is symmetric, $T = T_A = T_B$, and if
%\begin{equation} \label{eq:oblique_and_vertic_result2}
%  %\frac{b}{c}>\frac{1-T}{\alpha T}
%	c < b \cdot \frac{\alpha T}{1-T},
%\end{equation}
%then fixation of the cooperator phenotype $A$ is locally stable.
%The same condition was given in Corollary 1.1, \autoref{eq:equal_transmission}.

%We now turn to the general case where $T_A \neq T_B$. 
We apply \autoref{eq:horizontal}, \autoref{eq:nextgen_adults_slimpify}, and \autoref{eq:nextgen_parents_simplified} to obtain the function $f(\hat{p})$, see Appendix~\autoref{sec:appendixB}:
